---
title: 'Assignment 6: Data Scraping'
author: 'Leo Zhang'
output: pdf_document
geometry: margin = 1cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A06_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory


```{r 1. Project Setup, message = FALSE, warning = FALSE}
#1 
knitr::opts_chunk$set(tidy = TRUE, tidy.opts = list(width.cutoff = 20))
library(rvest)
library(tidyverse)
getwd()
```


2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Set the year to 2024
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)


```{r set.the.scraping.website, message = FALSE, warning = FALSE}
#2 
website <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')
glimpse(website)
```


3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * Contact Person
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "Reginald Hicks", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".


```{r scrape.the.data, message = FALSE, warning = FALSE}
#3 
introduce_first <- website %>% html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()
introduce_first
introduce_second <- website %>% html_nodes('div+ table tr:nth-child(4) td:nth-child(2)') %>% html_text()
introduce_second
introduce_third <- website %>% html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text()
introduce_third
introduce_fourth <- website %>% html_nodes('th~ td+ td') %>% html_text()
introduce_fourth
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.


```{r create.a.dataframe.from.scraped.data, message = FALSE, warning = FALSE}
#4
report_first <- data.frame('Water_System_Name' = rep(introduce_first, times = 12), 'Contact_Person' = rep(introduce_second, times = 12), 'Ownership' = rep(introduce_third, times = 12), 'Maximum_Day_Use' = as.numeric(introduce_fourth), 'Year' = rep(2024, times = 12), 'Month' = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12))
report_first <- report_first %>% mutate(Date = ym(paste(Year, '-', Month)))
str(report_first)
#5 
p1 <- ggplot(data = report_first, aes(x = Date, y = Maximum_Day_Use)) + geom_line(linewidth = 1, color = '#0096FF') + geom_smooth(method = 'loess', se = FALSE, linetype = 'dashed', linewidth = 1, color = '#F52757') + theme_gray() + labs(title = 'Maximum Day Use Of Water', subtitle = 'Data In 2024, From Durham', x = 'month', y = '(mgd)')
p1
```


6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs, "PWSID" and "year", that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function, message = FALSE, warning = FALSE}
#6. 
scrape <- function(year, facility){
  target <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', facility, '&year=', year))
  tag_first <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  tag_second <- 'div+ table tr:nth-child(4) td:nth-child(2)'
  tag_third <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  tag_fourth <- 'th~ td+ td'
  universal_first <- target %>% html_nodes(tag_first) %>% html_text()
  universal_second <- target %>% html_nodes(tag_second) %>% html_text()
  universal_third <- target %>% html_nodes(tag_third) %>% html_text()
  universal_fourth <- target %>% html_nodes(tag_fourth) %>% html_text()
  report_all <- data.frame('Water_System_Name' = rep(universal_first, times = 12), 'Contact_Person' = rep(universal_second, times = 12), 'Ownership' = rep(universal_third, times = 12), 'Maximum_Day_Use' = as.numeric(universal_fourth), 'Year' = rep(year, times = 12), 'Month' = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12)) %>% mutate(Date = ym(paste(Year, '-', Month)))
  return(report_all)
}
```


7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 


```{r fetch.and.plot.Durham.2020.data, message = FALSE, warning = FALSE}
#7
report_second <- scrape(2020, '03-32-010')
p2 <- ggplot(data = report_second, aes(x = Date, y = Maximum_Day_Use)) + geom_line(linewidth = 1, color = '#0096FF') + geom_smooth(method = 'loess', se = FALSE, linetype = 'dashed', linewidth = 1, color = '#F52757') + theme_gray() + labs(title = 'Maximum Day Use Of Water', subtitle = 'Data In 2020, From Durham', x = 'month', y = '(mgd)')
p2
str(report_second)
```


8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.


```{r fetch.and.plot.Cary.2020.data, message = FALSE, warning = FALSE}
#8 
report_third <- scrape(2020, '03-92-020')
comparison_first <- bind_rows(report_second, report_third)
p3 <- ggplot(data = comparison_first, aes(x = Date, y = Maximum_Day_Use, color = Water_System_Name)) + geom_line(linewidth = 1) + geom_smooth(method = 'loess', se = FALSE, linetype = 'dashed', linewidth = 0.75) + scale_color_manual(values = c('#0096FF', '#F52757')) + theme_gray() + labs(title = 'Maximum Day Use Of Water', subtitle = 'Data In 2020, From Cary And Durham', x = 'month', y = '(mgd)')
p3
```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "06_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs (where one of the two inputs does not change). Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.


```{r fetch.and.plot.Cary.multiyear.data, message = FALSE, warning = FALSE}
#9 
history <- c(2018, 2019, 2020, 2021, 2022, 2023)
location <- '03-92-020'
comparison_second <- map2(history, location, scrape) %>% bind_rows()
p4 <- ggplot(data = comparison_second, aes(x = Date, y = Maximum_Day_Use)) + geom_line(linewidth = 1, color = '#0096FF') + geom_smooth(method = 'loess', se = FALSE, linetype = 'dashed', linewidth = 1, color = '#F52757') + theme_gray() + labs(title = 'Maximum Day Use Of Water', subtitle = 'Data Between 2018 And 2023, From Cary', x = 'year', y = '(mgd)')
p4
```


Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?


> Answer: In general, the water use during summer is more than that during winter. Cary might witness increasing water use over time. However, this trend is not significant.

