---
title: 'Assignment 3: Data Exploration'
author: 'Leo Zhang'
date: '2026-02-03'
output: pdf_document
geometry: margin = 2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change ‘Student Name’ on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  [NEW] Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to Canvas.

8.  Initial here to acknowledge that you did not use AI in completing this assignment, except where expressly allowed: LZ

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks in your code chunks.

**TIP**: If your code fails to knit, check:
 * That no `install.packages()` or `View()` commands exist in your code. 
 * That you are not displaying the entire contents of a large dataframe in your code.

---

## Set up your R session

1.  Load necessary packages (tidyverse, here), check your current working directory and import two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets ‘Neonics’ and ‘Litter’, respectively. 

**Be sure to**:
* Use the `here()` package in specifying the paths to your datasets
* Include the appropriate subcommand to read in character based columns as factors


```{r specify_packages, message = FALSE, warning = FALSE}
# Load two necessary packages.
library(here)
library(tidyverse)
```


```{r import_data, message = FALSE, warning = FALSE}
# Introduce two dataframes.
getwd()
Neonics <- read.csv(file = here('./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv'), stringsAsFactors = TRUE)
Litter <- read.csv(file = here('./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv'), stringsAsFactors = TRUE)
```


## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. (AI is allowed here, but put answers in your own words.)

> Answer: Neonicotinoids can kill insects indiscriminately, and honeybees have encountered considerable threats. In addition, neonicotinoids can penetrate into plant cells and remain in water, which means that their presence in the natural environment is extensive. The chronic risk of neonicotinoids to humans is also possible.


3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. (AI is allowed here, but put answers in your own words.)

> Answer: Forest litters are associated with the wellness of surrounding ecosystems. They may reflect biodiversity, climate change, or decomposition patterns.


4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. The litter sampling is targeted to take place in numerous 40mx40m plots.
 2. Ground traps are sampled once per year.
 3. No single site should ever exceed 3440 data product instances in a given calendar year.


## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r basic_information, message = FALSE, warning = FALSE}
# Determine the dimension of 'Neonics'. There are 4623 rows and 30 columns.
str(Neonics)
```

6.  Using the `summary` function on the 'Effect' column, determine the most common effects that are studied.  [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r effect_classification, message = FALSE, warning = FALSE}
# List the distribution of 'Effect' column in an ascending sequence.
sort(summary(Neonics$Effect))
```


Question: Which two effects stand out as the most studied? Can you guess why these effects might specifically be of interest?
> Answer: Population and mortality are the most studied. They might be vital effects that an insect undergoes. They might be relatively easier to observe.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name).[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r species_classification, message = FALSE, warning = FALSE}
# List the distribution of 'Species.Common.Name' column in a descending sequence.
sort(summary(Neonics$Species.Common.Name, maxsum = 4623), decreasing = TRUE)
```


Question: What do these species have in common? Why might they be of interest over other insects? 
> Answer: Honey bee, parasitic wasp, buff tailed bumblebee, Carniolan honey bee, bumble bee, and Italian honeybee are the six species. They carry out significant economic benefits, so scholars might be interested in pursuing their research.


8.  The `Conc.1..Author` column, which lists the concentration of the neonicitoid dose, should include numeric values. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r verify_concentration, message = FALSE, warning = FALSE}
# Figure out the status of 'Conc.1..Author.' column.
class(Neonics$Conc.1..Author.)
```

> Answer: It is factor. The values, containing integers, divisions, or 'NR', seem to be messy and inconsistent.


## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r publication_simple, message = FALSE, warning = FALSE}
# Derive a line plot regarding the frequency of 'Publication.Year' column.
p1 <- ggplot(data = Neonics, aes(x = Publication.Year)) + geom_freqpoly(bins = 40, size = 1, color = '#000000') + labs(title = 'Number Of Studies Each Year', subtitle = 'Data From 4623 Observations', x = 'year', y = '')
p1
```


10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r publication_complex, message = FALSE, warning = FALSE}
# Determine the number of unique values in 'Test.Location' column.
unique(Neonics$Test.Location)
# Derive a differentiated line plot regarding the frequency of 'Publication.Year' column. A four-color palette is adopted.
p2 <- ggplot(data = Neonics, aes(x = Publication.Year, color = Test.Location)) + geom_freqpoly(bins = 20, size = 0.75) + scale_color_manual(values = c('#0096FF', '#E30B5C', '#2E8B57', '#FFD700')) + labs(title = 'Number Of Studies Conducted At Different Locations Each Year', subtitle = 'Data From 4623 Observations', x = 'year', y = '')
p2
```


Interpret this graph. What are the most common test locations, and do they differ over time?
> Answer: Lab is the most common test location. Most lab tests took place after 2010, but most field natural tests took place before 2010.


11. Create a bar graph of Endpoint counts. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r endpoint_simple, message = FALSE, warning = FALSE}
# Derive a bar plot regarding the frequency of 'Endpoint' column.
p3 <- ggplot(data = Neonics, aes(x = Endpoint)) + geom_bar(fill = '#0096FF') + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + labs(title = 'Number Of Studies Classified By Endpoint', subtitle = 'Data From 4623 Observations', x = 'endpoint', y = '')
p3
```


What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix (p.721) for more information. 
> Answer: NOEL and LOEL. 

---

## Explore your data (Litter)

12. Determine the class of `collectDate`. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r clean_date, message = FALSE, warning = FALSE}
# The first attempt shows that 'collectDate' column contains factor values.
class(Litter$collectDate)
# Change the format for 'collectDate' column. The second attempt shows that 'collectDate' column contains date values.
Litter$collectDate <- ymd(Litter$collectDate)
class(Litter$collectDate)
# There are two dates for collection: 2018-08-02, 2018-08-30.
unique(Litter$collectDate)
```


13. Using the `unique` function, list the different `plotIDs` sampled at Niwot Ridge. 

```{r verify_location, message = FALSE, warning = FALSE}
# Discover the number of unique values in 'plotID' column.
unique(Litter$plotID)
summary(Litter$plotID)
```


How is the information obtained from `unique` different from that obtained from `summary`?
> Answer: 'Summary' function shows all unique values and the frequency of each unique value. However, 'unique' function only shows all unique values.


14. Create a bar graph of `functionalGroup` counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r tissue_simple, message = FALSE, warning = FALSE}
# Derive a bar plot regarding the frequency of 'functionalGroup' column.
p4 <- ggplot(data = Litter, aes(x = functionalGroup)) + geom_bar(fill = '#0096FF') + labs(title = 'Number Of Records Classified By Functional Group', subtitle = 'Data From 188 Observations', x = 'functional group', y = '')
p4
```


15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of `dryMass` by `functionalGroup`.

```{r mass_complex, message = FALSE, warning = FALSE}
# Derive a box plot regarding the distribution of 'dryMass' column. The grouping factor is 'functionalGroup'.
p5 <- ggplot(data = Litter, aes(x = functionalGroup, y = dryMass)) + geom_boxplot(color = '#0096FF') + labs(title = 'Distribution Of Dry Mass Grouped By Functional Group', subtitle = 'Data From 188 Observations', x = 'functional group', y = 'dry mass')
p5
# Derive a violin plot regarding the distribution of 'dryMass' column. The grouping factor is 'functionalGroup'.
p6 <- ggplot(data = Litter, aes(x = functionalGroup, y = dryMass)) + geom_violin(color = '#000000') + labs(title = 'Distribution Of Dry Mass Grouped By Functional Group', subtitle = 'Data From 188 Observations', x = 'functional group', y = 'dry mass')
p6
```


Why is the boxplot a more effective visualization option than the violin plot in this case?
> Answer: The sample size is too small for a violin plot to present clear information.


What type(s) of litter tend to have the highest biomass at these sites?
> Answer: Needles tend to have the highest biomass.

